TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-06-10 ManuscriptTe.Rmd
MS updates

The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure 1). Profiles can be fit as a single smooth function or segments.
```{r, echo=FALSE}
#replace infinity with N
t.seq1[which(!is.finite(t.seq1))]=NA

@@

*Figure 1. Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()].*

@@

```{r, include=FALSE}
#cut nas
dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass","Ugrass")]
dat= na.omit(dat)
```

@@

#T_a	and T_g are air and surface temperatures (C), respectively; u	is wind speed (m/s); 
#H is total (direct + diffuse) solar radiation flux (W/m^2); 
#K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
#psi is solar zenith angle (degrees); l	is grasshopper length (m);
#Acondfact is the proportion of the grasshopper surface area in contact with the ground;
#abs is absorptivity of grasshopper to solar radiation (proportion);
#r_g is substrate solar reflectivity (proportion).
```

@@

We aggregate the data and plot average conditions across sites (Figure 2). The simplifying assumptions in the biophysical model implementation reduce alignment with the observed temperatures of grasshoppers, but the divergence of grasshopper temperatures from air temperatures at 2m and, to a lesser extent, air temperatures at grasshopper height demonstrate the importance of implementing biophysical models.
```{r, echo=FALSE, warning=FALSE}
dat$TeGates=TeGates-273.15

@@

*Figure 2. Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C). Microclimate models indicate that the air temperatures experienced by grasshoppers near the ground (Ta grasshopper height) is generally above air temperatures (Ta) and close to surface temperatures (Ts). Body temperatures estimated by a grasshopper specific biophysical model (Te) are fairly similar to but sometimes less than the body temperatures of a physical grasshopper model implanted with a thermistor (Te observed). The general biophysical models that makes simplifying assumptions (Te Gates) tends to overestimate body temperatures.*

>>>>

The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure S1). Profiles can be fit as a single smooth function or segments.
```{r, echo=FALSE}
#replace infinity with N
t.seq1[which(!is.finite(t.seq1))]=NA

@@

*Figure S1. Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()]. The points indicate empirical points used for parameterization.*

@@

```{r, include=FALSE}
#cut NAs
dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass","Ugrass")]
dat= na.omit(dat)
```

@@

#T_a	and T_g are air and surface temperatures (C), respectively; u	is wind speed (m/s); 
#H is total (direct + diffuse) solar radiation flux (W/m^2);
#K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
#psi is solar zenith angle (degrees); l	is grasshopper length (m);
#Acondfact is the proportion of the grasshopper surface area in contact with the ground;
#abs is absorptivity of grasshopper to solar radiation (proportion); abs=0.9 corresponds to dark colors.
#r_g is substrate solar reflectivity (proportion).
```

@@

We aggregate the data and plot average conditions across sites (Figure S2). The simplifying assumptions in the biophysical model implementation reduce alignment with the observed temperatures of grasshoppers, but the divergence of grasshopper temperatures from air temperatures at 2m and, to a lesser extent, air temperatures at grasshopper height demonstrate the importance of implementing biophysical models.
```{r, echo=FALSE, warning=FALSE}
dat$TeGates=TeGates-273.15

@@

*Figure S2. Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C). Microclimate models indicate that the air temperatures experienced by grasshoppers near the ground (Ta grasshopper height) is generally above air temperatures (Ta) and close to surface temperatures (Ts). Body temperatures estimated by a grasshopper specific biophysical model (Te) are fairly similar to but sometimes less than the body temperatures of a physical grasshopper model implanted with a thermistor (Te observed). The general biophysical models that makes simplifying assumptions (Te Gates) tends to overestimate body temperatures.*

### 2022-06-10 Manuscript_full.Rmd
MS updates

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

1.  Much understanding of organimsal responses to climate change and
    variability relies on the assumption that body temperatures are
    equal to temporally averaged air temperatures high above the ground.
    However, most organisms live near the ground and acute exposure to
    solar and thermal radiation and thermal extremes can substantially
    elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the
    organismal and environonmental parameters needed for biophysical
    ecology.
3.  TrenchR focuses on simple and modular functions so users can create
    transparent and flexible models for biophysical applications.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. We further this aim through a series of educational
    modules that introduce the field of biophysical ecology.

@@

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of
these tools has not kept pace with research on climate change responses.
Adoption of biophysical approaches is expanding with recognition of the
importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr] and the increased dissemination of computational
tools.

@@

**Figure 1** A) Microclimate models scale air and soil temperatures and windspeeds from sensor to organismal heights. B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral]. Grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.

The TrenchR package is intended to complement the NicheMapR package,
which includes sophisticated microclimate [@kearney2017nichemapr] and
biophysical [@kearney2020nichemapr] models. Although the Fortran source
code was recently released, the complexity of NicheMapR functions can
make it difficult to understand and modify the source code. Many of the
approaches are similar between the packages, but TrenchR provides simple
functions to aid the understanding and transparency of biophysical
approaches. The TrenchR functions are modular and easily adapt to a
variety of organisms and research questions. Although many of the
TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We omit plant specific biophysical
models since they are the focus of several R packages [tealeaves:
[@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. Some energy
budget calculation components of biophysical models are also available
in the ThermImage package [@tattersall2017]. Microclimate models, with
an emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles, which describe the angles of incident sunlight),
facilitate modeling diurnal temperature variation, and estimate
biologically relevant aggregate metrics such as degree days available
for development.

@@

**Specific biophysical functions** provide biophysical models that have
been built for particular organisms based on their physical properties,
behavior, environment, and life history. We currently provide published
biophysical models for lizards, salamanders, butterflies, grasshoppers,
limpets, mussels, and snails. Most models predict operative
environmental temperatures, which are the equilibrium body temperatures
of organisms with specified physical properties in a specific
microclimate and assume no heat exchange via metabolism or evaporation
[@bakken1992measurement].

@@

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is
also incorporated in the *Using energy balances to estimate body
temperatures* vignette. We start by generating environmental inputs.
Using these inputs, we estimate the energy budget with component
functions. Finally, we use an integrated biophysical model to estimate
operative environmental temperatures. See Appendix S1 for another
example examining a time series of microclimate data and applying and
testing the biophysical models.

We will use the energy budget to estimate body temperature, $T_b$, which
can depart dramatically from the air temperatures due to energy exchange
with the environment. Heat is gained from absorbed solar and thermal
radiation and from metabolic reactions. Heat is lost through the
organism's emission of radiation and the evaporation of water. The
organism exchanges heat with the surrounding air or water via convection
and with the ground via conduction. The balance of these heat exchanges
(omitting metabolism and evaporation, which are often negligible for
ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an
estimate of $T_b$ and the package functions refer to $T_b$ for
simplicity.

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2 meters) reports that the daily air
temperature varies from a minimum of 10°C to a maximum of 25°C and the
wind speed averages 1 m/s. The soil surface temperature varies from a
minimum of 15°C to a maximum of 30°C. We assume that atmospheric
transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$

@@

We first prepare the environmental data for analysis. We will estimate
hourly air and soil temperatures and radiation using a function
describing diurnal temperature variation. We start by estimating the day
of year and the timing of sunrise and sunset:

@@

We then calculate hourly air and soil temperatures based on daily
minimum and maximum temperatures. We select the sine-exponential model
for air temperature and the sine model for surface temperature:

@@

We use microclimate models to scale air temperature and wind speed from
weather station height (2m) to lizard height (0.02m). We assume a
surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and
determines the turbulence of airflow. We implement a neutral air
temperature profile, an air temperature profile forced by wind speed,
and a neutral wind speed profile.

@@

We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10 gram
*Sceloporus* lizard with solar absorptivity $a = 0.9$
[@gates1980biophysical]. We will initially assume $T_b = T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

@@

# Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxon = "lizard_surface")

# Estimate the heat transfer coefficient using a spherical approximation
H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxon = "lizard")

@@

TrenchR is intended to promote understanding of how organisms interact
with their environment and consequences for physiology, energetics,
behavior, and demography. Our example implementation highlights the
importance of considering organismal body temperatures, rather than air
temperatures, when examining thermal stress and other responses to
environmental variability and change. TrenchR currently focuses on heat
balances but may be expanded to include water balances. Simple functions
can be combined as needed to produce comprehensive and transparent
models for biophysical ecology and evolution. The resultant models are
likely to be sufficiently detailed for many applications, but users are
referred to NicheMapR for more detailed biophysical models
[@kearney2017nichemapr; @kearney2020nichemapr]. We focus on models that
predict equilibrium conditions for simplicity. Such models generally do
not present computational challenges so our models are not optimized for
computational efficiency. Making classic biophysical ecology techniques
more accessible will allow researchers to take advantage of rapidly
accumulating data on environmental conditions and organismal traits to
understand and predict ecological and evolutionary responses.
Considering how organisms experience their environment is central to
understanding responses to variable and changing environments
[@suggitt2018extinction].

>>>>

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

1.  Much understanding of organimsal responses to climate change and
    variability relies on the assumption that body temperatures are
    equal to temporally averaged air temperatures high above the ground.
    However, most organisms live near the ground and acute exposure to
    solar and thermal radiation and thermal extremes can substantially
    elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the
    organismal and environmental parameters needed for biophysical
    ecology.
3.  TrenchR focuses on simple and modular functions so users can create
    transparent and flexible models for biophysical applications.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. We further this aim through a series of educational
    modules that introduce the field of biophysical ecology.

@@

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of
these tools has not kept pace with research on climate change responses.
In recent decades, adoption of biophysical approaches is expanding with recognition of the
importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr] and the increased dissemination of computational
tools.

@@

**Figure 1** A) Microclimate models scale air and soil temperatures and windspeeds from sensor to organismal heights. B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral]. A thermal image depicts how grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.

The TrenchR package is intended to complement the NicheMapR package,
which includes sophisticated microclimate [@kearney2017nichemapr] and
biophysical [@kearney2020nichemapr] models. Although the Fortran source
code was recently released, the complexity of NicheMapR functions can
make it difficult to understand and modify the source code. Many of the
approaches are similar between the packages, but TrenchR provides simple
functions to aid the understanding and transparency of biophysical
approaches. Some variable definitions and units are lacking in NicheMapR. The TrenchR functions are modular and easily adapt to a
variety of organisms and research questions. Although many of the
TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We omit plant specific biophysical
models since they are the focus of several R packages [tealeaves:
[@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. Some energy
budget calculation components of biophysical models are also available
in the ThermImage package [@tattersall2017]. Microclimate models, with
an emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles, which describe the angles of incident sunlight),
facilitate modeling diurnal temperature variation, and estimate
biologically relevant aggregate metrics such as degree days available
for oranismal development.

@@

**Specific biophysical functions** provide biophysical models that have
been built for particular organisms based on their physical properties,
behavior, environment, and life history. We currently provide published
biophysical models for lizards, salamanders, butterflies, grasshoppers,
limpets, mussels, and snails. Most models predict operative
environmental temperatures, which are the equilibrium body temperatures
of organisms with specified physical properties in a specific
microclimate and assume no heat exchange via metabolism or evaporation
[@bakken1992measurement]. However, we also present an analytical function for humid operative temperature that incorporates the effects of evaporative cooling on operative temperatures for wet-skinned ectotherms, such as salamanders.

@@

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is
also incorporated in the *Using energy balances to estimate body
temperatures* vignette. We start by generating environmental inputs.
Using these inputs, we estimate the energy budget with component
functions. Finally, we use an integrated biophysical model to estimate
operative environmental temperatures. See Appendix S1 for another
example examining a time series of microclimate data and applying and
testing the biophysical models.
**Flow cart of functions?

We will use the energy budget to estimate body temperature, $T_b$, which
can depart dramatically from the air temperatures due to energy exchange
with the environment. Heat is gained from absorbed solar and thermal
radiation and from metabolic reactions. Heat is lost through the
organism's emission of radiation and the evaporation of water. The
organism exchanges heat with the surrounding air or water via convection
and with the ground via conduction. The balance of these heat exchanges
(omitting metabolism and evaporation, which are often negligible for
ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an
estimate of $T_b$ and the package functions refer to $T_b$ for
simplicity. Our examplar exstimation of $T_b$ assumes thermal equilibrium.

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2 meters) reports that the daily air
temperature varies from a minimum of 10°C to a maximum of 25°C and the
wind speed averages 1 m/s. The soil surface temperature varies from a
minimum of 15°C to a maximum of 30°C. We assume that atmospheric
transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.

@@

At the first stage, we prepare the environmental data for analysis. We will estimate
hourly air and soil temperatures and radiation using a function
describing diurnal temperature variation. We start by estimating the day
of year and the timing of sunrise and sunset:

@@

We then calculate hourly air and soil surface temperatures based on daily
minimum and maximum temperatures. We select the sine-exponential model
for air temperature and the sine model for surface temperature:

@@

At the second stage, we use microclimate models to scale air temperature and wind speed from
weather station height (2m) to lizard height (0.02m). We assume a
surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and
determines the turbulence of airflow. We implement a neutral air
temperature profile driven by density differences, an air temperature profile forced by wind speed,
and a neutral wind speed profile.

@@

Finally, we will use our microclimates estimates to solve the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10 gram
*Sceloporus* lizard with solar absorptivity $a = 0.9$
[@gates1980biophysical]. We will initially assume $T_b = T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

@@

# Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")

# Estimate the heat transfer coefficient using a spherical approximation
H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")

@@

TrenchR is intended to promote understanding of how organisms interact
with their environment and consequences for physiology, energetics,
behavior, and demography. Our example implementation highlights the
importance of considering organismal body temperatures, rather than air
temperatures, when examining thermal stress and other responses to
environmental variability and change. TrenchR currently focuses on heat
balances but may be expanded to include water balances. Simple functions
can be combined as needed to produce comprehensive and transparent
models for biophysical ecology and evolution. The resultant models are
likely to be sufficiently detailed for many applications, but users are
referred to NicheMapR for more detailed biophysical models
[@kearney2017nichemapr; @kearney2020nichemapr]. We focus on models that
predict equilibrium conditions for simplicity (that is, steady-state conditions). Such models generally do not present computational challenges so our models are not optimized for computational efficiency. Making classic biophysical ecology techniques
more accessible will allow researchers to take advantage of rapidly
accumulating data on environmental conditions and organismal traits to
understand and predict ecological and evolutionary responses.
Considering how organisms experience their environment is central to
understanding responses to variable and changing environments
[@suggitt2018extinction].
